Capon Beamforming Applied on Moving Objects - An Analysis of Local Shift-Invariance

### 2013-10-26 HRUIofMovingObjects/jrnl_article.tex
Moving objects paper: Added new LSV plots

An important property of an ultrasound imaging system is its local shift-invariance. Local shift-invariance means that an object’s appearance should be preserved when the object, or the ultrasound probe, is moved slightly. This local movement invariance is what makes ultrasound imaging feasible in general, however it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood. In this paper we analyze the shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.

Capon beamforming has been shown to provide increased lateral resolution and contrast in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. However, it is also known for its suboptimal performance if the sample size available for statistical estimation is limited, if correlated signals are present, and if the assumed steering vector does not match the signal propagation vector\cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer using either spatial smoothing \cite{Shan1985}, diagonal loading \cite{JianLi2003} or steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}. The first two, together with temporal smoothing, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}, and both trades resolution for robustness. The latter has not been applied, even though it is well known that the Capon beamformer exhibits severe signal-nulling if the steering vector does not match the signal steering vector exactly. This self-nulling actually contribute a lot to the method's super resolution capability in passive-narrowband applications.

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer (called the $\vec{k}_3$ processor), and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor): 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution. Impinging on the array is a plane wave with a propagation vector orthogonal to the array surface normal with no spatially white noise present. Hence, the interference-noise matrix $\sigma_0^2\mat{Q}$ is the identity. We have also added the effect of typical sub-array averaging and diagonal loading \cite{Asen} to the Capon beamformer output. The figure shows that when the DAS beamformer is used, only a 3.92 dB scallop loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array. However if the Capon beamformer is used with the beam density sufficient for DAS the scallop loss could be as large as $31$ dB.

@@

In previous literature on Capon beamforming for medical ultrasound imaging this effect has been passed over in silence (in single frames) by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or ignorance. In this paper we show that when imaging a moving point scatterer with ultrasound and the Capon beamformer with the same beam spacing sufficient for DAS the same scallop loss is observed, manifested as oscillating point scatteres. The question is then, what can we do in order to improve on the situation.

@@

The sample covariance matrix can be estimated in the following way \cite{Synnevag2009}:
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
\end{align}
where $K = M-L+1$ typical is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$. Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness, 
\begin{align}
\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
\end{align}

@@

Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computational demanding, and has for long limited the methods accessibility. Following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, this has now changed \cite{Asen}.

@@

\subsection{Sampling and Shift-Invariance}

The lateral sampling of an ultrasound image is given by the aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) combined \cite{Hergum2007} and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the media and $\lambda$ is the wavelength. The required Nyquist beam spacing is then:
\begin{align}
\Delta < \frac{\lambda}{D_{tx} + D_{rx}}. \label{eq:resolution}
\end{align}
If this requirement is not fulfilled, aliasing effects will appear in the image and the system will not be lateral shift invariant.

In order to visualize the local lateral shift-invariance of a given imaging system, we will use lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer. For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked in order to produce an image. The system is said to be local lateral shift-invariant if the plot shows straight diagonals with close to constant amplitude.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_no_undersampling.eps}\label{fig:das_shiftinvariant}}
\hfill{}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_undersampled_2x.eps}\label{fig:das_shiftvariant}}
}
\caption{...}
\label{fig:das}
\end{figure*}

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_normal.eps}\label{fig:capon}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_normal.eps}
}
}
\caption{...}
\label{fig:image_capon}
\end{figure*}

@@

\section{Simulation Setup}

>>>>

An important property of an ultrasound imaging system is its local shift-invariance. Local shift-invariance means that an object’s appearance should be preserved when the object, or the ultrasound probe, is moved slightly. This invariance for local movements is what makes ultrasound imaging feasible in general, however it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood. In this paper we analyze the shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.

Capon beamforming has been shown to provide increased lateral resolution and contrast in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. The method accomplish this by adjusting the array weights based on element-to-element covariance. Hence, the method yields data dependent weights. Capon beamforming is however also known for its suboptimal performance if the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets (SVUS) \cite{Lorenz2005, Rubsamen2013}. The first two, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trades resolution for robustness and temporal smoothing helps to preserve speckle statistics \cite{Synnevag2007a}. SVUS has not been applied, even though it is well known that the Capon beamformer exhibits severe signal-nulling if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly. This self-nulling actually contribute a lot to the method's super resolution capability in passive-narrowband applications.

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer (called the $\vec{k}_3$ processor), and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor): 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution. Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface normal with no spatially white noise present. Hence, the interference-noise matrix $\sigma_0^2\mat{Q}$ is the identity. We have also added the effect of typical subarray averaging and diagonal loading \cite{Asen} to the Capon beamformer output. The figure shows that when the DAS beamformer is used, only a 3.92 dB scallop loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array. However if the Capon beamformer is used with the beam density sufficient for DAS the scallop loss could be as large as $31$ dB.

@@

In previous literature on Capon beamforming for medical ultrasound imaging this effect has been passed over in silence (in single frames) by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or by ignorance. In this paper we show that when imaging a moving point scatterer with ultrasound and the Capon beamformer with the same beam spacing sufficient for DAS the same scallop loss is observed, manifested as oscillating point scatteres. The question is then, what can we do in order to improve on the situation.

@@

The sample covariance matrix can be estimated for an active system in the following way \cite{Synnevag2009}:
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
\end{align}
where $K = M-L+1$ typically is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$. Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness, 
\begin{align}
\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
\end{align}

@@

Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}.

@@

\subsection{Lateral Sampling and Shift-Invariance}

The lateral sampling of an ultrasound image is given by the aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) combined \cite{Hergum2007} and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the media and $\lambda_c$ is the wavelength of the center frequency. The required Nyquist beam spacing is then:
\begin{align}
\Delta < \frac{\lambda}{D_{tx} + D_{rx}}. \label{eq:resolution}
\end{align}
If this requirement is not fulfilled, aliasing effects will appear in the image and the system will not be lateral shift-invariant.

In order to visualize the local lateral shift-invariance of a given imaging system, we will use lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer. For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked in order to produce an image. The image is displayed as a contour plot with a contour on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot. The system is said to be local lateral shift-invariant if the contours are straight diagonals, hence the diagonals have constant amplitude.

To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus $f(x)=x$ (P-MAE), and the negative MAE of the amplitude variation for the same set of points (A-MAE). The first measure quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal. The maximum amplitude deviation along the max-point trace is therefore also highly interesting. We present all these numbers together with the LSV-plots.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
\hfill{}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
}
\caption{...}
\label{fig:das}
\end{figure*}

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2.eps}\label{fig:capon}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_2.eps}
}
}
\caption{...}
\label{fig:image_capon}
\end{figure*}

@@

\section{In-vitro Images}
Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.

\section{In-Vivo Images}
Investigate if phase rotation improves contrast in in-vivo images of the heart.
